The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending

### 2015-03-29 draft.Rnw
Added Capital One data. Described Feature Selection. Refocused abstract

\title{The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer\footnote{R code available at \url{github.com/sshleifer/ProsperThesis}}\\ Yale University}
\date{\today}
\maketitle 
\begin{abstract}
This paper documents violations of market efficiency in Peer to Peer Lending, using data from Prosper.com, America's first Peer-to-Peer Lending Marketplace.
Peer to Peer lending is the practice of individual and institutional investors lending money to borrowers over the internet, without the assistance of a traditional financial intermediary.
Theories of Market Efficiency predict that asset ``prices should fully reflect publicly available information." 
Contrary to this prediction, strategies using publicly available information can
earn superior risk-adjusted returns, when compared with the other loans.

This violation of market efficiency, and poor investor returns in general, are
discussed in the midst of two market design changes by Prosper that guided, and
then forced investors towards setting more efficient interest rates. Such
guidance and eventual price control, as opposed to the unconstrained market
design, where interest rates were set by auction, is associated with improved
investor returns and more efficient markets. The aforementioned strategies can
only earn excess returns in the first two regimes, before Prosper began directly
setting interest rates.
\end{abstract}
\newpage

\section{Introduction}
Historically, there are two major perspectives of efficient market prices. Fama
((1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a
strategy using publicly available information should not be able to earn excess
risk-adjusted returns. Risk models, such as CAPM, imply that risky assets should
have positive expected returns and that riskier assets should have higher
returns than less risky assets. The information aggregation perspective
associated with (Hayek, 1945) and (Stiglitz and Grossman, 1980), further
predicts that many individuals, each using their own information, collectively
arrive at a more efficient price than a central agent who sets prices, since the
agent cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

In the P2P Lending context, the market's price is the interest rate associated with a loan. This paper studies the question of whether interest rates on Prosper.com are set efficiently. (Fama, 1970)'s semi-strong EMH predicts that if interest rates were set efficiently, then...
\begin{enumerate}
\item Loans would have positive expected returns. Investors would be unwilling to lend to Borrowers if they expected to lose money. \footnote{Assuming a Risk-Free Rate of 0}
\item Riskier loans would have higher expected returns. \footcite{sharpe}
\item It is not possible to do better than the average investor using publicly available information. \footcite{fama}
\end{enumerate}
I test hypothesis 1 by measuring the historical returns of loans up to a certain date, and then asking whether loans with negative expected returns get funded at that date.
I test hypothesis 2 by comparing the returns of high risk loans to low risk loans.
I test hypothesis 3 by using publicly available information to select loans with a higher than average expected return.

Prosper.com data provide some additional valuable insights into investor behavior. As it happens, over the period I study, there are three interest rate determination regimes:
\begin{enumerate}
\item Early Auction Regime (April 19, 2006-October 15, 2008): Rates are set by investor auction. Investors have large amounts of data, but Prosper does not guide them meaningfully.
\item Late Auction Regime: (July 1, 2009-December 20, 2010): Rates are set by investor auction, with Prosper offering valuable guidance in the form of more complex credit grading.
\item Prosper-rate Regime: (December 21, 2010 - August 3, 2011): Prosper sets rates using its more complex credit grading algorithm.
\end{enumerate}

The early auction regime is closest to Hayek's vision of the free market, since interest rates are determined freely by investors through auctions.
I find that it is precisely in this free-market regime that the three
predictions of the EMH are contradicted.
In the two later regimes, when Prosper first provides much more guidance to investors, and then actually sets interest rates, investor returns improve dramatically.
The evidence across regimes is broadly inconsistent with the view that local knowledge and the wisdom of the crowd improves price discovery.
To the contrary, it appears that a "centralized' Prosper mechanism of either guiding or setting interest rates actually leads to fewer predictable mispricings: as Prosper becomes more involved in the rate-setting process, the excess returns available using publicly available information approach 0, while aggregate returns exhibit higher mean and lower variance.

@@

In this spirit, I do not claim that poor investor decision making is the only
factor causing poor returns in the Early auction period, or that the impact of
the policy changes can be accurately isolated by my methodologies.
The contradictions with efficiency hypotheses 1 and 2 can certainly be partially
explained by the same bad luck that explains poor returns on all risky assets
during economically turbulent times.
I merely argue that returns in the early auction period were likely made worse
by poor investor decisions, as evidenced by the predictable mispricings that can
be exploited, as they are in Section 5, the core of my market inefficiency
argument, which finds that over a somewhat longer period of time, simple loan selection strategies can use publicly available information to earn higher returns than the benchmark.\footnote{Benchmark calculated by assuming \$1 investment in every available loan.}

\subsection{Background}
Peer-to-peer lending (``P2PL")  is the practice of lending money to borrowers over the internet, without going through a traditional financial intermediary. \footnote{\url{en.Wikipedia.org/wiki/Peer-to-peer_lending}} From April 2006 until December 2010, interest rates were set by lenders, who competed for the lowest rate in a reverse auction model.\newline
In October 2008, Prosper was shut down by the SEC, and re-opened after a 9 month quiet period with more stringent under-writing standards, and greater guidance for lenders . Importantly, all the information about borrower creditworthiness that is used in this paper was available to investors when they bid on loans.
Since December 2010, the beginning of \textit{Prosper-rate} regime, rates have been pre-set by Prosper on the basis of a Borrower's Credit history.

@@

I do not discount or annualize ROIs, because in the period of low interest rates, the difference between default and repayment is large enough that a few percentage points don't matter: loans that default average -46\% ROI, loans that repay average + 24\% ROI.\footnote{Elaborate?}

@@

For credit buckets above 600, the relationship between credit score and default rate appears to be fairly linear, for all three regimes, suggesting that borrowers should be sensitive to credit scores when setting interest rates, since they are predictive of default.\footnote{I have a chart (suggested by Lapo) that shows what rate investors would have needed to charge to break even. It shows, as expected, that investors in early regime would have needed to charge much higher rates, especially at risky credit grades to break even. Should I put it in?}
Early auction regime investors, then, were not sufficiently sensitive to the hard credit variables that are predictive of default, especially when evaluating risky, low-credit borrowers, suggesting that the poor returns of the early auction regime were not caused only by the time period, but also by poor investor decisions.\footnote{NEED MORE HERE} We now look at the more informative grading algorithm Prosper began using after the shutdown, which made investors more sensitive to hard credit variables.

@@

In the next sub section, I examine Prosper's other interventions, which ban low-credit borrowers, and show that these interventions also similarly improved investor returns.
\subsection{Impact of Low-Credit Borrower Ban}
<<csvtime,cache=FALSE,fig.height=3, fig.cap="This figure presents average estimated credit score of loans originated each month. The average credit score increases from 653 in the early auction regime to 714 in the late auction regime, before returning to 705 in the Prosper-rate regime. Grey clouds indicate one standard deviation.">>=
ggplot(mmeans, aes(x=mgroup, y=mcs, color=regime)) +
  geom_point(aes(size=numLoans))+
  geom_smooth(aes(ymax=mcs.up, ymin=mcs.down),stat='identity') +
  #geom_line() +
  theme_bw() +
  labs(title='Borrower Credit Scores Over Time', size='Number of Loans',x='Origination Month', y='Mean Borrower Credit Score')
##TODO(ADD ERROR BARS)
@
We now turn to Prosper's policy changes, which restrict the types of borrowers who investors are allowed to lend to. An auction perspective predicts that the crowd of lenders should be able to effectively screen all candidates, not just low risk candidates. On multiple occasions, however, Prosper has intervened to restrict the borrower pool to high credit borrowers, and thereby improved investor returns. Figure \ref{fig:csvtime} shows the mean borrower credit score for each month in the sample, with the grey clouds showing one standard deviation bounds. Overall, for the three samples in full, the average borrower credit scores increases from 653.2 before the shutdown to 714.2 after the shutdown, then dips to 705 for the Prosper-rate period.
For the first 10 months of Prosper's history, investors choose to lend to borrowers with an average estimated credit score of 609. Prosper's policy change, on March 1, 2007, that prevented loans to ``low-credit borrowers," defined as those with credit score below 520, showed an immediate positive impact on investor decision making and returns.

Prosper's reasoning for the change was to protect investors: by March 1, 2007, 14.9\% of loans to low-credit borrowers had already defaulted, more than triple the 4.6\% of loans to other borrowers that had already defaulted. \footnote{the 10 \% gap in default rates is even more striking when the average age of the two groups of loans is considered. On March 1, 2007, the loans to low-credit borrowers averaged 100 days since origination, while loans to other borrowers averaged 122 days since origination. 14.9\% of low credit borrowers defaulted over an average time since origination of 100 days, whereas only 4.6\% of other borrowers defaulted over an average time since origination of 122 days.} Theoretically, however, it is more difficult to understand why Prosper needed to step in. From the EMH perspective, theory predicts that investors would have reacted to the news of the defaults on low credit loans by decreasing their exposure to low credit loans, or increasing the interest rates they bid on them. However, the median interest rate on low-credit, actually decreased from 24.23\% for the month of July, 2006 to 23.4\% for February, 2007, the month before the intervention. Similarly, exposure to low credit loans did not substantially decrease. On December 1, 2006, 13.3\% of these bad loans had already, after an average of only 3 monthly payments, gone into default, as compared to 3.7\% of other loans. Yet over the month of December 2006, loans to borrowers with credit score below 520 accounted for 22\% of orginations, the highest proportion in the site's history.  The wisdom of the crowd theory expects market participants to adjust their beliefs to new information more rapidly and flexibly than a central planner. The data, however, show that investors continued to make avoidably bad decisions until Prosper disallowed them.

@@

Although Figure \ref{fig:ratevscore} shows that investors charge higher rates to riskier borrowers, the high risk, low-reward data pattern of the early auction regime, shown in Figure \ref{fig:roivscore} suggests that this preference for more credit-worthy borrowers is not extreme enough, since the extra interest charged to riskier borrowers does not, in aggregate, make up for the difference in default likelihood. This suggests that the variables given to investors by Prosper could be better used to assess borrower creditworthiness, and select loans with the lower likelihood of default. In this section, I use a regression model (that a naive investor might be able to use to choose loans) to show that a simple model to assess borrower creditworthiness can select loans, out of sample, that earn higher returns than the loans funded by investors. First I carefully discuss the methodology, which sets up the linear regressions that (1) can be tested by the mean ROI of the loans they choose out of sample, and thus will be penalized for overfitting, (2) are trained on data that would not have been available at the time of the decision (3) uses automated feature selection, so that I, with the benefit of hindsight, don't choose which variables will be included.
\subsection{Methodology}
\subsubsection{Cross Validation Procedure}
<<crossValexp, cache=FALSE>>=
cval = data.frame('Sample'= c('Auction-Early.train', 'Auction-Early.test','Auction-Late-.train','Auction-Late.test',
                             'Prosper-rate.train','Prosper-rate.test'),
                 'Loans'= c(nrow(train),nrow(test),nrow(tr2),nrow(te2),nrow(tr3),nrow(te3)),
                 'Start' =  c(min(train$Date),min(test$Date),min(tr2$Date),min(te2$Date),min(tr3$Date),min(te3$Date)),
                 'End' =  c(max(train$Date),max(test$Date),max(tr2$Date),max(te2$Date),max(tr3$Date),max(te3$Date))) %>%
  transform(Days = as.character(as.numeric(difftime(End, Start))), Start = as.character(Start), End=as.character(End))
xtable(cval, label='cval', caption='This table documents number of loans and dates of the 6 samples used for training and testing regression strategies.')
@
The regime discontinuity splits the data into early auction, late auction and Prosper-rate regimes. To test our predictions out of sample, we split each of these thirds in half. For those familiar with cross validation, regressions are trained on the first half of loans originated under each regime, and tested on the second half. Table \ref{cval} describes the size and periods of the six sub samples. For those less familiar with cross validation, I now describe the six sub samples in words.
\begin{enumerate}
\item \Sexpr{nrow(train)} loans originated between April 21, 2006 and  October 2nd, 2007, are used to generate coefficients for the early auction regime regressions.
\item \Sexpr{nrow(test)} loans originated after October 2nd, 2007, but before October 16, 2008, are used to test  the predictions of the early auction regime regressions. \footnote{Between October 10, 2008 and July 20, 2009, no loans are originated, because Prosper is in the SEC-imposed Quiet Period}
\item \Sexpr{nrow(tr2)} loans originated between July 20, 2009 and  April 26, 2010, are used to generate coefficients for the late auction regime regressions. 
\item \Sexpr{nrow(te2)} loans originated after April 26, 2010, but before December 17, 2010, are used to test the predictions of the late auction regime regressions.
\item \Sexpr{nrow(tr3)} loans originated between Dec 20,2010 and April 28,2011  are used to generate coefficients for the Prosper-rate regime regressions.
\item \Sexpr{nrow(te3)} loans originated between April 29,2010 and August 3rd 2011 are used to generate coefficients for the Prosper-rate regime regressions.
\end{enumerate}

@@

\subsubsection{Out of Sample Loan Selection}
After training linear models to predict Known-Default, we use the coefficients to predict known-default as a proxy for real default, for each possible loan in the testing sample, and chooses the 10\% of loans from each credit grade with the lowest predicted default likelihood. \footnote{Loans are selected from each credit grade to ensure that strategies in the later periods can actually earn positive returns. Grade A loans in later periods return less than average.}
\subsubsection{Summary of Procedure}
For each regime \ldots
\begin{enumerate}
\item divide each regime into training and testing samples of equal size, with all loans in the training sample originated before the first loan in the test sample was originated.
\item create variable Known-Default $= (0 || 1)$ for each loan, depending on whether an investor would have known that the loan defaulted at the end of the training sample.
\item Start with 45 Features, select 8 that best predict Known-default.\footcite{leaps}\footnote{Feature selection was performed by selecting the best combination of 8 features out of 45, based on in-sample performance (scoring with Bayesian Information Criterion).}
\item Run Logistic Regression to predict Known-Default using training period data and selected variables
\item Using trained coefficients, generate predicted Known-Default values for loans in the test period.
\item Select the 10\% of Loans from each credit grade with the lowest predicted Known-default probability.
\item Compare ROI of selected loans against the  benchmark of a strategy that chooses every loan.
\end{enumerate}

\subsubsection{Assumptions}
Two assumptions underly the out-of sample loan selection technique which selects from the pool of loans that were funded by Prosper's investors, at the rate they received. First, in the auction regime, by sucessfully bidding on a loan, investors may decrease the interest rate. This analysis assumes that a strategy does not move the interest rates of the loans it selects. Second, a strategy never chooses to fund loans that will end up unfunded, since we cannot observe their ROIs.
\subsection{In Sample Results}
<<runRegs,cache=FALSE>>=
form1.sc = makeForm(bigReg1.sc,'def ~')
form2.sc = makeForm(bigReg2.sc,'def ~')
form3.sc = makeForm(bigReg3.sc,'def ~')

@@

In summary, over the inital two regimes where investors are involved in the process of setting interest rates, via auction, it is possible to use publicly available information to earn excess returns. In the final regime, where the central planner sets the interest rate, it is much more difficult, if not impossible, to earn excess returns using publicly available information.

@@

\section{Conclusion (unfinished)}
 EMH predicts that (1) risky assets should have positive expected returns, (2) riskier assets should have higher expected return than less risky assets and (3) it should be impossible to use publicly available information to earn excess returns. The previous sections document evidence that rejects all three of these predictions, as well as evidence that the inconsitencies with EMH were diminished and then removed as Prosper took more control of the rate-setting process. I conclude by offering alternative framings of the central planner as a market guiding, rather than disrupting force.
\newpage
\printbibliography
\end{document}

>>>>

\title{The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending}
\author{Samuel Shleifer\footnote{R code available at
\url{github.com/sshleifer/ProsperThesis}}\\ Senior Essay in Economics}
\date{Yale University, \today}
\maketitle 
\begin{abstract}
This essay analyzes the effect of information gathering costs on financial market
efficiency, using data from Prosper.com, America's first Peer-to-Peer Lending Marketplace.
Peer-to-peer lending is the practice of individual and institutional investors
lending money to borrowers over the internet without the assistance of a
traditional financial intermediary. Between 2006 and 2011, Prosper underwent two
major policy changes that guided and then forced investors to set higher
interest rates, and reduce their exposure to subprime borrowers.

During the policy regime before either policy intervention (2006-2008), investors set
interest rates by auction, made avoidably bad mistakes, and earned poor returns.
The data from the first regime contradicts three Efficient Markets
Hypothesis predictions: that (1) risky assets should earn positive expected returns,
(2) riskier assets should earn higher returns than less risky assets, and, most
conclusively, (3) ``prices should fully reflect publicly available information."

Controlling for macroeconomic differences between the time periods of each policy regime
reveals that contradictions (1) and (2), which disappeared after Prosper began
guiding investors towards setting higher interest rates, are partly the
consequence of a turbulent macroeconomic period, but that better investor
decisions, or Prosper setting interest rates from the beginning, would have
weakened the blow of the financial crisis. Contradiction (3), that strategies
using publicly available information can earn superior risk-adjusted returns,
which holds until Prosper begins directly setting rates, suggesting that 
centralized interest rate setting may lead to a more efficient market than the
unconstrained interaction of supply and demand, especially when information
gathering costs are high.
\end{abstract}
\newpage

\section{Introduction}
Historically, there are two major perspectives of efficient market prices. Fama
((1970)'s semi-strong efficient markets hypothesis ("EMH") predicts that a
strategy using publicly available information should not be able to earn excess
risk-adjusted returns. Risk models such as CAPM imply that risky assets should
have positive expected returns and that riskier assets should have higher
returns than less risky assets.\footcite{sharpe} The information aggregation perspective
associated with Hayek (1945) and Stiglitz and Grossman (1980), further
predicts that many individuals, each using their own information, collectively
arrive at a more efficient price than a central agent who sets prices, since the
agent cannot access all the information available to the crowd.\footcite{stiglitz}\footcite{hayek}

In the P2P Lending context, the market's price is the interest rate associated with a loan. This paper studies the question of whether interest rates on Prosper.com are set efficiently. (Fama, 1970)'s semi-strong EMH predicts that if interest rates were set efficiently, then...
\begin{enumerate}
\item Loans would have positive expected returns. Investors would be unwilling
  to lend to Borrowers if they expected to lose money. \footnote{Assuming a
  Risk-Free Rate of 0.}
\item Riskier loans would have higher expected returns. \footcite{sharpe}
\item It is not possible to do better than the average investor using publicly available information. \footcite{fama}
\end{enumerate}
I test hypothesis 1 by measuring the historical returns of loans up to a certain date, and then asking whether loans with negative expected returns get funded at that date.
I test hypothesis 2 by comparing the returns of high risk loans to low risk loans.
I test hypothesis 3 by using publicly available information to select loans with a higher than average expected return.

Prosper.com data provide some additional valuable insights into investor behavior. As it happens, over the period I study, there are three interest rate determination regimes:
\begin{enumerate}
\item Early Auction Regime (April 19, 2006-October 15, 2008): Rates are set by investor auction. Investors have large amounts of data, but Prosper does not guide them meaningfully.
\item Late Auction Regime: (July 1, 2009-December 20, 2010): Rates are set by investor auction, with Prosper offering valuable guidance in the form of more complex credit grading.
\item Prosper-rate Regime: (December 21, 2010 - August 3, 2011): Prosper sets rates using its more complex credit grading algorithm.
\end{enumerate}
Additionally, for regimes (2) and (3), subprime borrowers are not allowed to
apply for Prosper loans.

The early auction regime is closest to Hayek's vision of the free market, since
interest rates are determined freely by investors through auctions, with few
limitations on which borrowers they can invest in.
I find that it is precisely in this free-market regime that the three
predictions of the EMH are contradicted.
In the two later regimes, when Prosper first provides much more guidance to investors, and then actually sets interest rates, investor returns improve dramatically.
The evidence across regimes is broadly inconsistent with the view that local knowledge and the wisdom of the crowd improves price discovery.
To the contrary, it appears that a "centralized' Prosper mechanism of either guiding or setting interest rates actually leads to fewer predictable mispricings: as Prosper becomes more involved in the rate-setting process, the excess returns available using publicly available information approach 0, while aggregate returns exhibit higher mean and lower variance.

@@

In this spirit, I do not claim that poor investor decision making is the only
factor causing poor returns in the Early auction period, or that the impact of
the policy changes can be accurately isolated by my methodologies.
The contradictions with efficiency hypotheses 1 and 2 can certainly be partially
explained by the same bad luck that explains poor returns on all risky assets
during economically turbulent times.
I merely argue that returns in the early auction period were likely made worse
by poor investor decisions, as evidenced by the predictable mispricings that can
be exploited, as they are in Section 5, the core of my market inefficiency
argument, which finds that over a somewhat longer period of time, simple loan
selection strategies can use publicly available information to earn higher
returns than the benchmark.\footnote{Benchmark is calculated by assuming \$1 investment in every available loan.}

\subsection{Background}
Peer-to-peer lending (``P2PL")  is the practice of lending money to borrowers
over the internet, without going through a traditional financial
intermediary.\footcite{p2pwiki} From April 2006 until December 2010, interest
rates were set by lenders\footnote{For the purposes of this essay, lender and
investor are synonyms.}, who competed for the lowest rate in a reverse auction
model.

In October 2008, Prosper was shut down by the SEC, and re-opened in July 2009
after a 9 month quiet period with more stringent under-writing standards, and
greater guidance for lenders. Importantly, all the information about borrower
creditworthiness that is used in this paper is similar to what was available to
investors when they bid on loans. Since December 2010, the beginning of the
\textit{Prosper-rate} regime, rates have been determined by Prosper, on the
basis of a borrower's credit history.

@@

I do not discount or annualize ROIs, because in the period of low interest
rates, the difference between default and repayment is large enough that a few
percentage points don't matter: loans that default average -46\% ROI, loans that
repay average + 24\% ROI. Furthermore, my ROI results are very similar as Freedman
and Jin's, who use discounting.\footcite{freedman}

@@

For credit buckets above 600, the relationship between credit score and default rate appears to be fairly linear, for all three regimes, suggesting that borrowers should be sensitive to credit scores when setting interest rates, since they are predictive of default.\footnote{I have a chart (suggested by Lapo) that shows what rate investors would have needed to charge to break even. It shows, as expected, that investors in early regime would have needed to charge much higher rates, especially at risky credit grades to break even. Should I put it in?}
Early auction regime investors, then, were not sufficiently sensitive to the hard credit variables that are predictive of default, especially when evaluating risky, low-credit borrowers, suggesting that the poor returns of the early auction regime were not caused only by the time period, but also by poor investor decisions. We now look at the more informative grading algorithm Prosper began using after the shutdown, which made investors more sensitive to hard credit variables.

@@

In the next sub section, I examine Prosper's other interventions, which ban low-credit borrowers, and show that these interventions also similarly improved investor returns.
\subsection{Impact of Subprime Borrower Ban}
<<csvtime,cache=FALSE,fig.height=3, fig.cap="This figure presents average estimated credit score of loans originated each month. The average credit score (using bucket midpoint) increases from 653 in the early auction regime to 714 in the late auction regime, before returning to 705 in the Prosper-rate regime. Grey clouds indicate one standard deviation.">>=
ggplot(mmeans, aes(x=mgroup, y=mcs, color=regime)) +
  geom_point(aes(size=numLoans))+
  geom_smooth(aes(ymax=mcs.up, ymin=mcs.down),stat='identity') +
  #geom_line() +
  theme_bw() +
  labs(title='Borrower Credit Scores Over Time', size='Number of Loans',x='Origination Month', y='Mean Borrower Credit Score')
##TODO(ADD ERROR BARS)
@
We now turn to Prosper's policy changes, which restrict the types of borrowers who investors are allowed to lend to. An auction perspective predicts that the crowd of lenders should be able to effectively screen all candidates, not just low risk candidates. On multiple occasions, however, Prosper has intervened to restrict the borrower pool to high credit borrowers, and thereby improved investor returns. Figure \ref{fig:csvtime} shows the mean borrower credit score for each month in the sample, with the grey clouds showing one standard deviation bounds. Overall, for the three samples in full, the average borrower credit scores increases from 653.2 before the shutdown to 714.2 after the shutdown, then dips to 705 for the Prosper-rate period.
For the first 10 months of Prosper's history, investors choose to lend to borrowers with an average estimated credit score of 609. Prosper's policy change, on March 1, 2007, that prevented loans to ``low-credit borrowers," defined as those with credit score below 520, showed an immediate positive impact on investor decision making and returns.

Prosper's reasoning for the change was to protect investors: by March 1, 2007, 14.9\% of loans to low-credit borrowers had already defaulted, more than triple the 4.6\% of loans to other borrowers that had already defaulted. \footnote{the 10\% gap in default rates is even more striking when the average age of the two groups of loans is considered. On March 1, 2007, the loans to low-credit borrowers averaged 100 days since origination, while loans to other borrowers averaged 122 days since origination. 14.9\% of low credit borrowers defaulted over an average time since origination of 100 days, whereas only 4.6\% of other borrowers defaulted over an average time since origination of 122 days.} Theoretically, however, it is more difficult to understand why Prosper needed to step in. From the EMH perspective, theory predicts that investors would have reacted to the news of the defaults on low credit loans by decreasing their exposure to low credit loans, or increasing the interest rates they bid on them. However, the median interest rate on low-credit, actually decreased from 24.23\% for the month of July, 2006 to 23.4\% for February, 2007, the month before the intervention. Similarly, exposure to low credit loans did not substantially decrease. On December 1, 2006, 13.3\% of these bad loans had already, after an average of only 3 monthly payments, gone into default, as compared to 3.7\% of other loans. Yet over the month of December 2006, loans to borrowers with credit score below 520 accounted for 22\% of orginations, the highest proportion in the site's history.  The wisdom of the crowd theory expects market participants to adjust their beliefs to new information more rapidly and flexibly than Prosper. The data, however, show that investors continued to make avoidably bad decisions until Prosper disallowed them.

@@

Although Figure \ref{fig:ratevscore} shows that investors charge higher rates to riskier borrowers, the high risk, low-reward data pattern of the early auction regime, shown in Figure \ref{fig:roivscore} suggests that this preference for more credit-worthy borrowers is not extreme enough, since the extra interest charged to riskier borrowers does not, in aggregate, make up for the difference in default likelihood. This suggests that the variables given to investors by Prosper could be better used to assess borrower creditworthiness, and select loans with the lower likelihood of default. In this section, I use a regression model (that a naive investor might be able to use to choose loans) to show that a simple model to assess borrower creditworthiness can select loans, out of sample, that earn higher returns than the loans funded by investors. First I carefully discuss the methodology, which sets up the linear regressions that (1) can be tested by the mean ROI of the loans they choose out of sample, and thus will be penalized for overfitting, (2) are trained on data that would not have been available at the time of the decision (3) uses automated feature selection, so that I, with the benefit of hindsight, don't choose which variables will be included.
\subsection{Methodology}

\subsubsection{Cross Validation Procedure}
<<crossValexp, cache=FALSE>>=
cval = data.frame('Sample'= c('Auction-Early.train', 'Auction-Early.test','Auction-Late-.train','Auction-Late.test',
                             'Prosper-rate.train','Prosper-rate.test'),
                 'Loans'= c(nrow(train),nrow(test),nrow(tr2),nrow(te2),nrow(tr3),nrow(te3)),
                 'Start' =  c(min(train$Date),min(test$Date),min(tr2$Date),min(te2$Date),min(tr3$Date),min(te3$Date)),
                 'End' =  c(max(train$Date),max(test$Date),max(tr2$Date),max(te2$Date),max(tr3$Date),max(te3$Date))) %>%
  transform(Days = as.character(as.numeric(difftime(End, Start))), Start = as.character(Start), End=as.character(End))
xtable(cval, label='cval', caption='This table documents number of loans and dates of the 6 samples used for training and testing regression strategies.')
@
The regime discontinuity splits the data into early auction, late auction and Prosper-rate regimes. To test our predictions out of sample, we split each of these thirds in half. For those familiar with cross validation, regressions are trained on the first half of loans originated under each regime, and tested on the second half. Table \ref{cval} describes the size and periods of the six sub samples. For those less familiar with cross validation, I now describe the six sub samples in words.
\begin{enumerate}
\item \Sexpr{nrow(train)} loans originated between April 21, 2006 and  October 2nd, 2007, are used to generate coefficients for the early auction regime regressions.
\item \Sexpr{nrow(test)} loans originated after October 2nd, 2007, but before October 16, 2008, are used to test  the predictions of the early auction regime regressions. \footnote{Between October 10, 2008 and July 20, 2009, no loans are originated, because Prosper is in the SEC-imposed Quiet Period}
\item \Sexpr{nrow(tr2)} loans originated between July 20, 2009 and  April 26, 2010, are used to generate coefficients for the late auction regime regressions. 
\item \Sexpr{nrow(te2)} loans originated after April 26, 2010, but before December 17, 2010, are used to test the predictions of the late auction regime regressions.
\item \Sexpr{nrow(tr3)} loans originated between Dec 20,2010 and April 28,2011  are used to generate coefficients for the Prosper-rate regime regressions.
\item \Sexpr{nrow(te3)} loans originated between April 29,2010 and August 3rd 2011 are used to generate coefficients for the Prosper-rate regime regressions.
\end{enumerate}

@@

\subsubsection{Feature Selection}
The following procedure is executed on the training data for each of the three regimes to determine the 8 features\footnote{I use the terms 'feature' and 'Independent Variable' interchangeably.'} to use in the regression. It is automated by the leaps R package\footcite{leaps}, so I cannot choose features that I know to be powerful.
\begin{enumerate}
\item Start with 45 features, chosen by me because they don't have missing data.\footnote{A complete list of features can be found on \url{github.com/sshleifer/ProsperThesis}}.
\item Scale the features to have 0 mean and unit variance, by subtracting the mean from each observation and then dividing by the standard deviation.
\item For each possible 8-feature subset of the original features, run a linear regression predicting Known-Default, and recover the Bayesian Information Criterion.\footnote{For more on BIC, \url{http://en.wikipedia.org/wiki/Bayesian_information_criterion}}
\item Choose the subset of features with the highest BIC. These 8 features will then be used in the training regressions.
\end{enumerate}

\subsubsection{Out of Sample Loan Selection}
After training linear models to predict Known-Default, we use the coefficients to predict known-default as a proxy for real default, for each possible loan in the testing sample, and chooses the 10\% of loans from each credit grade with the lowest predicted default likelihood. \footnote{Loans are selected from each credit grade to ensure that strategies in the later periods can actually earn positive returns. Grade A loans in later periods return less than average.}
\subsubsection{Summary of Procedure}
For each regime \ldots
\begin{enumerate}
\item divide the loans into training and testing samples of equal size, with all loans in the training sample originated before the first loan in the test sample was originated.
\item create variable Known-Default $= (0 || 1)$ for each loan, depending on whether an investor would have known that the loan defaulted at the end of the training sample.
\item Start with 45 Features, select the 8 feature subset that best predicts Known-default.\footcite{leaps}
\item Run Linear Regression to predict Known-Default using training period data and selected variables.
\item Using trained coefficients, generate predicted Known-Default values for loans in the test period.
\item Select the 10\% of Loans from each credit grade with the lowest predicted Known-default probability.
\item Compare ROI of selected loans against the  benchmark of a strategy that chooses every loan.
\end{enumerate}

\subsubsection{Assumptions}
Two assumptions underly the out-of sample loan selection technique which selects from the pool of loans that were funded by Prosper's investors, at the rate they received. First, in the auction regime, by sucessfully bidding on a loan, investors may decrease the interest rate. This analysis assumes that a strategy does not move the interest rates of the loans it selects. Second, a strategy never chooses to fund loans that will end up unfunded, since we cannot observe their ROIs.

\subsection{In Sample Results}
<<runRegs,cache=FALSE>>=
form1.sc = makeForm(bigReg1.sc,'def ~')
form2.sc = makeForm(bigReg2.sc,'def ~')
form3.sc = makeForm(bigReg3.sc,'def ~')

@@

In summary, over the inital two regimes where investors are involved in the process of setting interest rates, via auction, it is possible to use publicly available information to earn excess returns. \footnote{Note that excess returns are as compared to other Prosper loans, not other asset classes like treasuries.} In the final regime, where the central planner sets the interest rate, it is much more difficult, if not impossible, to earn excess returns using publicly available information.

@@

\subsection{Comparison to Traditional Loans (Unfinished)}

<<macro,cache-FALSE>>=
#<600 11.84
#601-660 17.84
#661-720 30.48
#Over 720 39.16
nd$year = year(nd$mgroup)
grouper<- function(x){
  return(ifelse(x < 660, .2969,
                      ifelse(x < 720, .3048,
                                    ifelse(x > 720, .3969, .2969))))
}
nd = transform(nd, cg = grouper(CreditScoreEst))
nd.ts = group_by(nd, regime, year, cg) %>% 
  summarise(def.rate = mean(defaulted), n = length(defaulted)) %>% filter(n > 103)
nd.ts= mutate(ungroup(nd.ts), year = as.numeric(year), def.part = def.rate*cg)
ts.adj = summarise(group_by(nd.ts,regime,year), def.adj = sum(def.part), n=sum((n)))

mac = read.csv('Data/comet.csv',col.names=c('year','def.rate3'))
# ts.raw = group_by(nd, regime, year) %>% 
#   summarise(def.rate = mean(defaulted), n = length(defaulted)) %>% filter(n > 300)
ts = inner_join(mac, ts.adj, by='year') %>% arrange(year) 
colnames(ts) = c('Origination','CapitalOne.default','Regime','Prosper.default','N.Prosper')
xtable(ts, label='macro', caption= 'This table documents the default rate of Prosper Loans (Prosper.Default) with the default rate in the first three years of consumer credit loans securitized by Capital One (CapitalOne.Default). Additionally, the regime and number of Prosper loans originated in each period are reported.')

xx = filter(read.csv('Data/c1fico.csv'), Credit.Score != '')
xtable(xx, label='macro.samp', caption='The Portion of borrowers from each credit score group in the Capital One data.')
@
By comparing the performance of Prosper loans originated in each period to the performance of traditional consumer credit loans originated in each period, controlling for credit score as best we can\footnote{Method: Given the default rates of some consumer loans, and a table of the frequency of a bunch of credit scores, I created the same credit score groups in the Prosper data, took each group's mean default rate, and took the weighted average default rate for each year, with the group's representation in the Capital One borrower pool, shown in Table \ref{macro.samp} as the weights. Since the Capital One loans are of different maturities, I measured the portion of defaults within three years of origination.} gain a better understanding of the time period. The data, presented in Table \ref{macro} show that for all default rates are higher for Prosper loans than Capital One loans, but that the difference is much larger in the early auction regime. for the prosper data than the Capital One data, but the difference is most extreme for the early auction regime. There may again be flawed assumptions, including adverse selection of borrowers on the internet, other credit variables besides score, and the different maturities of the loans, but the data again suggest that early auction regime investors did a particularly bad job.\footnote{This may be in the right place. I spent a while looking for the ideal data on the web and this is the closest I came.}

\section{Conclusion (TODO)}
\begin{enumerate}
\item Restate results.
\item Central planner intervention may be useful when info gathering costs are high.
\item Future research: same experiment with known ROI. Better time controls? Look at consistent errors in mistakes borrowers made.
\end{enumerate}
\newpage
\printbibliography
\end{document}
